Why does preregistration increase the persuasiveness of evidence? A Bayesian rationalization.

### 2022-08-09 manuscript.Rmd

These issues arise from a fundamental confusion about the objective of preregistration.
Researchers are bound to be confused when they delegate an important scientific judgment to a simple decision rule, that only takes into account the existence of a preregistration.
Preregistering an inherently exploratory analysis (like testing dozens of relations) does not make it confirmatory, nor will a carefully conducted confirmatory study become exploratory if the researcher deviates from the preregistration in minor details.
Equating confirmatory research with preregistered research is only possible for studies that expect a low type-I error rate and will run/be analyzed without changes no matter whatin any case [@simmonsPreregistrationWhyHow2021; @bakkerEnsuringQualitySpecificity2020].
Under these conditions, the rule holds, but such restricted use makes preregistration a niche solution unable to match the greater problem of lacking replicability in psychology and elsewhere.

@@

Let us start by defining what we call expected epistemic value.
If researchers plan to conduct a study, they usually hope it will change their assessment of some theory's verisimilitude.
In other words, they hope to learn something from conducting the study.
The amount of knowledge researchers gain from a particular study concerning the verisimilitude of a specific theory is what we call epistemic value.
While researchers can not know what exactly they will learn from a study, they can form an expectation that helps them decide which study to conduct.
This expectation is what we term expected epistemic value.
To make our three arguments, we must assume three things about this estimation process and how it relates to what studies to conduct.

@@

The assumption of rationality can be connected to Bayesian reasoning and leads to our adoption of the framework.
Our rationale is as follows.
Researchers who decide to conduct a certain study are actually choosing a study to bet on.
They have to "place the bet" by conducting the study, therefore, invest resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study as a betting problem allows us to apply a "Dutch Book" argument [@christensenCleverBookiesCoherent1991].
This argument states that any better must follow the axioms of probability to avoid being "irrational," i.e., accepting bets that lead to sure losses.
Fully developing a Dutch book argument for this problem requires careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modelling what liberties researchers have in what studies to conduct.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are sufficient to derive the Bayes formula, on which we will heavily rely for our further arguments.
The argument is not sufficient, however, to warrant conceptualizing the kind of epistemic value we reason about in terms of probability; that remains a leap of faith.
Please note that our decision to adopt this aspect of the Bayesian philosophy of science does not imply anything about the statistical methods researchers use.
In fact, this conceptualization is purposefully reductionistic to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.

@@

Let us make some simplifying assumptions and define notation.
We restrict ourselves to evidence of a binary nature (either it exists or does not) since continuous evidence would lead to some quite involved derivations.
We denote the probability of a hypothesis before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
We can calculate the probability of the hypothesis after observing the evidence with the help of the Bayes formula:

@@

In short, we want to increase posterior probability ($P(H|E)$).
Increases in posterior probability ($P(H|E)$) are associated with increased epistemic value, for which we want to maximize the expectation.
So how can we increase posterior probability?
The Bayes formula yields three components that influence corroboration, namely $P(H)$, $P(E|H)$ and $P(E)$.
The first option leads us to the unsurprising conclusion that higher a priori probability ($P(H)$) leads to higher posterior probability ($P(H|E)$).
However, the prior probability of a hypothesis is nothing our study design can change.
The second option is similarly commonsensical; that is, an increase in $P(E|H)$ leads to higher posterior probability ($P(H|E)$).
$P(E|H)$ is the probability of obtaining evidence for a theory, when the theory holds.
We call this probability of detecting evidence given that the theory holds "detectability."
Consequently, researchers should ensure that their study design allows them to find evidence for their hypothesis, in case it is true.
When applied strictly within the bounds of null hypothesis testing, detectability is equivalent to power (or the inverse of type-II error rate).
However, while detectability is of great importance for study design, it is not directly relevant to the objective of preregistration.
Thus, $P(E)$ remains to be considered.
Since $P(E)$ is the denominator, increasing it will decrease the posterior probability: The more unlikely it is to observe evidence, the more it increases the probability of the hypothesis if we do observe it.
In other words, high risk, high reward.

@@

We already have noted that there is not much to do about prior probability ($P(H)$, and hence its counter probability $P(¬H)$), and that it is common sense to increase detectability ($P(E|H)$).
The real lever to pull is, therefore, $P(E|¬H)$.
This probability tells us how likely it is that we find evidence in favor of the theory when in fact, the theory is not true.
Its counter probability $P(¬E|¬H)= 1 - P(E|¬H)$ is what we call "theoretical risk", because it is the risk a theory takes on in predicting the occurrence of particular evidence in its favour.
We "borrow" the term from @meehlTheoreticalRisksTabular1978, though he has not assigned it to the probability $P(¬E|¬H)$.
In a response to the original paper, @kuklaClinicalStatisticalTheory1990 argued that the arguments layed out in @meehlTheoreticalRisksTabular1978 can be reconstructed in purely Bayesian framework.
However, while he did not name $P(¬E|¬H)$ but did suggest that @meehlTheoreticalRisksTabular1978 used the term "very strange coincidence" for a small $P(E|¬H)$ which would imply that $P(¬E|¬H)$ is theoretical risk.

@@

This sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove that a hypothesis is false ($P(¬E|¬H) = 0$, theoretical risk is zero), it cannot be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rational we apply here is able to reconstruct many commonly held views on riskiness and epistemic value.

@@

## Causes of uncertainty

As we noted, it is possible to quantify how statistical models affect the theoretical risk based on mathematical considerations and simulation.
However, other factors in the broader context of the study are much harder to quantify.
If one chooses to focus only on the contribution of statistical methods to theoretical risk, one is bound to overestimate it.
Take, for example, a t-test.
Under ideal circumstances (assumption of independence, normality of residuals, equal variance), it stays true to its type-I error rate.
However, researchers might do many very reasonable things in the broader context of the study that affect theoretical risk:
They might exclude outliers, choose to drop an item, enlarge their definition of the population to be sampled, translate their questionnaires, impute missing values, or any number of other things.
All of these decisions carry a small risk that they increase the likelihood of obtaining evidence despite the underlying hypothesis being false.
Even if the t-test itself perfectly maintains its type-I error rate, these factors must be added to $P(E|¬H)$ and, hence, be subtracted from theoretical risk.
While, in theory, these factors may leave theoretical risk unaffected or even increase it, we argue that this is not the case in practice.
Whether researchers want to or not, they continuously process information about how the study is going, except under strict blinding.
While one can hope that processing this information does not affect their decision making either way, this cannot be secured.
The only thing we can conclude with some certainty is that theoretical risk is not higher than what the statistical model guarantees without knowledge about the other factors at play.

@@

We chose this uniform distribution to capture our statement that a statistical model only guarantees an upper bound to theoretical risk (and since it is a probability the lower bound is 0) as it is the maximum entropy distribution under this assumption and thus conforms to our Bayesian framework [@giffinUpdatingProbabilitiesData2007].

@@

We hope to have persuaded you to accept two arguments: First, the theoretical risk is important for judging evidential support for theories.
Second, the theoretical risk is inherently uncertain, which diminishes the persuasiveness of the gathered evidence.
The third and last argument is that preregistrations reduce this uncertainty.

@@

Communicating clearly what authors did to arrive at the evidence they present is crucial for judging the theoretical risk they took.
Preregistrations are ideal to communicate just that, because any description after the fact is suspect to be incomplete.
For instance, the authors could have decided to exclude a number of analytic strategies they tried out.
That is not to say that every study that was not-preregistered was subjected to practices of p hacking.
The point is, that we cannot exclude this and innumerable possibilities and, hence, are left uncertain.
If the researchers have described what they intended to do beforehand, and then report that they did exactly that, the readers can be certain, that they have received a complete account of the situation.
They still might be uncertain about the actual theoretical risk the authors took, but to a much smaller extend than if the study would not have been preregistered.
Remaining sources of uncertainty might be unfamiliarity with statistical methods or experimental paradigms, the probability of plain error in analyses, etc.
In any case, a well written preregistration should aim to reduce the uncertainty about the theoretical risk and hence increase the persuasiveness of evidence.

@@

We started out with the observation that preregistrations do not always cleanly divide preregistration from confirmation.
Enforcing this distinction would mean using preregistrations as a means to increase the theoretical risk researchers take on.
No doubt this results in more trustworthy evidence and may even increase the effectiveness of scientific undertakings in general.
However, as the term implies, directly maximizing theoretical risk increases the likelihood that a study fails to deliver results favoring the theory.
As we showed, minimizing the uncertainty around the theoretical risk shares the beneficial properties without increasing the likelihood of a "failed" study.
Let us note that a study that does not result in the evidence a researcher is hoping for is still valuable for the scientific process.
Under the current incentive structure, however, many researchers have reason to avoid such outcomes and few, if any, have the privilege to operate outside of these incentives.

The advantage of this perspective that puts uncertainty about theoretical risk front and center is twofold.
First, aiming to reduce uncertainty is beneficial across a wide range of potential theoretical risks.
That is, researchers benefit from preregistering their study, regardless of whether the study is "exploratory" or not.
If researchers conduct a more exploratory study they can clearly communicate how exploratory they aim to be and their results can be judged accordingly.
Second, if researchers realize after the fact that it would be unwise to follow their preregistration in certain details, they may change their strategy.
Allowing researchers to deviate from the preregistration is controversial (XXX) because the authors might sift through all the possibilities and then select the most beneficial for their theory.
In its fullest consequence, this argument might result in the same uncertainty that we argued plagues for not-preregistered studies.
However, if the authors offer a convincing argument for their deviation, their readers might may judge it to be a reasonable explanation for the deviation.
In fact, using the here presented framework, we can distinguish reasonable from unreasonable deviations.
If a change increases detectability enough to offset the increase in uncertainty in theoretical risk, it is worthwhile.

>>>>

These issues arise from a fundamental confusion about the objective of preregistration.
Researchers are bound to be confused when they delegate an important scientific judgment to a simple decision rule, that only takes into account the existence of a preregistration.
Preregistering an inherently exploratory analysis (like testing dozens of relations) does not make it confirmatory, nor will a carefully conducted confirmatory study become exploratory if the researcher deviates from the preregistration in minor details.
Equating confirmatory research with preregistered research is only possible for studies that expect a low type-I error rate and will run/be analyzed without in any case [@simmonsPreregistrationWhyHow2021; @bakkerEnsuringQualitySpecificity2020].
Under these conditions, the rule holds, but such restricted use makes preregistration a niche solution unable to match the greater problem of lacking replicability in psychology and elsewhere.

@@

Let us start by defining what we call expected epistemic value.
If researchers plan to conduct a study, they usually hope it will change their assessment of some theory's verisimilitude.
In other words, they hope to learn something from conducting the study.
The amount of knowledge researchers gain from a particular study concerning the verisimilitude of a specific theory (which itself is an ontologocial concept) is what we call epistemic value.
Researchers cannot know what exactly they will learn from a study before they have run it.
However, they can develop, they can form an expectation that helps them decide which study to conduct in what manner.
This expectation is what we term expected epistemic value.
To make our three arguments, we must assume three things about this estimation process and how it relates to what studies (preregistered vs not preregistered) to conduct.

@@

The assumption of rationality can be connected to Bayesian reasoning and leads to our adoption of the framework.
Our rationale is as follows.
Researchers who decide to conduct a certain study are actually choosing a study to bet on.
They have to "place the bet" by conducting the study, therefore, invest resources and stand to gain epistemic value with some probability.
This conceptualization of choosing a study as a betting problem allows us to apply a "Dutch Book" argument [@christensenCleverBookiesCoherent1991].
This argument states that any better must follow the axioms of probability to avoid being "irrational," i.e., accepting bets that lead to sure losses.
Fully developing a Dutch book argument for this problem requires careful consideration of what kind of studies to include as possible bets, defining a conversion rate from the stakes to the reward, and modelling what liberties researchers have in what studies to conduct.
Without deliberating these concepts further, we find it persuasive that researchers should not violate the axioms of probability if they have some expectation about what they stand to gain with some likelihood from conducting a study.
The axioms of probability are sufficient to derive the Bayes formula, on which we will heavily rely for our further arguments.
The argument is not sufficient, however, to warrant conceptualizing the kind of epistemic value we reason about in terms of probability; that remains a leap of faith.
Please note that our decision to adopt this aspect of the Bayesian philosophy of science does not imply anything about the statistical methods researchers use.
In fact, this conceptualization is intentionally reductionistic to be compatible with a wide range of philosophies of science and statistical methods researchers might subscribe to.

@@

Let us make some simplifying assumptions and define notation.
We restrict ourselves to evidence of a binary nature (either it was observed or not) since continuous evidence would lead to some quite involved derivations.
We denote the probability of a hypothesis before observing evidence as $P(H)$ and its complement as $P(¬H) = 1 - P(H)$.
The probability of observing evidence under some hypothesis is $P(E|H)$.
We can calculate the probability of the hypothesis after observing the evidence with the help of the Bayes formula:

@@

In short, we want to increase posterior probability ($P(H|E)$).
Increases in posterior probability ($P(H|E)$) are associated with increased epistemic value, for which we want to maximize the expectation.
So how can we increase posterior probability?
The Bayes formula yields three components that influence corroboration, namely $P(H)$, $P(E|H)$ and $P(E)$.
The first option leads us to the unsurprising conclusion that higher a priori probability ($P(H)$) leads to higher posterior probability ($P(H|E)$).
If a hypothesis is more probable to begin with, the same increase in corroboration will lead to a higher probability than if the study was less probable to begin with.
However, the prior probability of a hypothesis is nothing our study design can change.
The second option is similarly commonsensical; that is, an increase in $P(E|H)$ leads to higher posterior probability ($P(H|E)$).
$P(E|H)$ is the probability of obtaining evidence for a theory, when the theory holds.
We call this probability of detecting evidence given that the theory holds "detectability."
Consequently, researchers should ensure that their study design allows them to find evidence for their hypothesis, in case it is true.
When applied strictly within the bounds of null hypothesis testing, detectability is equivalent to power (or the inverse of type-II error rate).
However, while detectability is of great importance for study design, it is not directly relevant to the objective of preregistration.
Thus, $P(E)$ remains to be considered.
Since $P(E)$ is the denominator, increasing it will decrease the posterior probability: The more unlikely it is to observe evidence, the more it increases the probability of the hypothesis if we do observe it.
In other words, high risk, high reward.

@@

We already have noted that there is not much to do about prior probability ($P(H)$, and hence its counter probability $P(¬H)$), and that it is common sense to increase detectability ($P(E|H)$).
The real lever to pull is, therefore, $P(E|¬H)$.
This probability tells us how likely it is that we find evidence in favor of the theory when in fact, the theory is not true.
Its counter probability $P(¬E|¬H)= 1 - P(E|¬H)$ is what we call "theoretical risk", because it is the risk a theory takes on in predicting the occurrence of particular evidence in its favour.
We "borrow" the term from @meehlTheoreticalRisksTabular1978, though he has not assigned it to the probability $P(¬E|¬H)$.
In a response to the original paper, @kuklaClinicalStatisticalTheory1990 argued that the arguments layed out in @meehlTheoreticalRisksTabular1978 can be reconstructed in purely Bayesian framework.
However, while he did not name $P(¬E|¬H)$ but did suggest that @meehlTheoreticalRisksTabular1978 used the term "very strange coincidence" for a small $P(E|¬H)$ which would imply that $P(¬E|¬H)$ can be related to or even equated to theoretical risk.

@@

The last statement sounds like a truism but is directly related to Popper's seminal criterion of demarcation.
He stated that if it is impossible to prove that a hypothesis is false ($P(¬E|¬H) = 0$, theoretical risk is zero), it cannot be considered a scientific hypothesis [@popperLogicScientificDiscovery2002, p. 18].
We note these relations to underline that the Bayesian rational we apply here is able to reconstruct many commonly held views on riskiness and epistemic value.

@@

## Sources of Uncertainty

As we noted, it is possible to quantify how statistical models affect the theoretical risk based on mathematical considerations and simulation.
However, other factors in the broader context of the study are much harder to quantify.
If one chooses to focus only on the contribution of statistical methods to theoretical risk, one is bound to overestimate it.
Take, for example, a t-test  of mean differences in two samples.
Under ideal circumstances (assumption of independence, normality of residuals, equal variance), it stays true to its type-I error rate.
However, researchers might do many very reasonable things in the broader context of the study that affect theoretical risk:
They might exclude outliers, choose to drop an item before computing a sum score, broaden, enlarge their definition of the population to be sampled, translate their questionnaires into a different language, impute missing values, or any number of other things.
All of these decisions carry a small risk that they increase the likelihood of obtaining evidence despite the underlying research hypothesis being false.
Even if the t-test itself perfectly maintains its type I error rate, these factors influence $P(E|¬H)$.
While, in theory, these factors may leave $P(E|¬H)$ unaffected or even decrease it, we argue that this is not the case in practice.
Whether researchers want to or not, they continuously process information about how the study is going, except under strict blinding.
While one can hope that processing this information does not affect their decision making either way, this cannot be secured.
We, therefore, conclude that statistical properties only guarantee a lower bound to the quantity we seek to minimize.
The only thing we can conclude with some certainty is that theoretical risk is not higher than what the statistical model guarantees without knowledge about the other factors at play.

@@

We chose this uniform distribution to capture our statement that a statistical model only guarantees an upper bound to theoretical risk (and since it is a probability the lower bound is 0) as it is the maximum entropy distribution under this assumption and conforms therefore to our Bayesian framework [@giffinUpdatingProbabilitiesData2007].

@@

We hope to have persuaded the reader to accept two arguments: First, the theoretical risk is important for judging evidential support for theories.
Second, the theoretical risk is inherently uncertain and the degree of uncertainty diminishes the persuasiveness of the gathered evidence.
The third and last argument is that preregistrations reduce this uncertainty.

@@

Communicating clearly how the authors gathered the data and consequently analyzed it to arrive at the evidence they present is crucial for judging the theoretical risk they took.
Preregistrations are ideal to communicate just that, because any description after the fact is suspect to be incomplete.
For instance, the authors could have decided to exclude a number of analytic strategies they tried out.
That is not to say that every study that was not-preregistered was subjected to practices of p hacking.
The point is, that we cannot exclude this and innumerable possibilities and, hence, are left uncertain.
This uncertainty is drastically reduced, if the researchers have described what they intended to do beforehand, and then report that they did exactly that.
In that case, the readers can be certain, that they have received a complete account of the situation.
They still might be uncertain about the actual theoretical risk the authors took, but to a much smaller extend than if the study would not have been preregistered.
Remaining sources of uncertainty might be unfamiliarity with statistical methods or experimental paradigms used, the probability of an implementation error in the statistical analyses, a bug in the software used for analysies, etc.
In any case a well written preregistration should aim to reduce the uncertainty about the theoretical risk and hence increase the persuasiveness of evidence.

@@

We started out with the observation that preregistrations do not always cleanly divide preregistration from confirmation.
Enforcing this distinction would mean using preregistrations as a means to increase the theoretical risk researchers take on.
No doubt this results in more trustworthy evidence and may even increase the effectiveness of scientific undertakings in general.
However, as argued earlier, directly maximizing theoretical risk increases the likelihood that a study fails to deliver results favoring the theory (a special instance of this problem is the well-known trade-off of statistical power and type-I error rate).
As we further argued, minimizing the uncertainty about theoretical risk (without necessarily increasing the theoretical risk itself) has benefits for the epistemic value, too, but without increasing the likelihood of producing a "failed" study.
Note that any single study that does not produce evidence in favor of a research hypothesis is still precious for the scientific process.
Under the current incentive structure, however, many researchers have reason to avoid such outcomes and few, if any, have the privilege to operate outside of these incentives.

In our view, the advantage of uncertainty reduction about theoretical risk front and center is two-fold.
First, aiming to reduce uncertainty is beneficial across a wide range of potential theoretical risks.
That is, researchers benefit from preregistering their study, regardless of whether or not the study is "exploratory".
If researchers conduct a more exploratory study they can clearly communicate how exploratory they aim to be and their results can be judged accordingly.
Second, if researchers realize after the fact that it would be unwise to follow their preregistration in certain details, they may change their strategy.
Allowing researchers to deviate from the preregistration is controversial (XXX) because the authors might sift through all the possibilities and then select the most beneficial for their theory.
In its fullest consequence, this argument might result in the same uncertainty that we argued plagues for not-preregistered studies.
However, if the authors offer a convincing argument for their deviation, their readers might may judge it to be a reasonable explanation for the deviation.
In fact, using the here presented framework, we can distinguish reasonable from unreasonable deviations.
If a change increases detectability enough to offset the increase in uncertainty in theoretical risk, it is worthwhile.
